\documentclass[12pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array}
\usepackage[table]{xcolor}
\usepackage{amsmath}

% Define the colors used in the CPE confusion table
\definecolor{TPgreen}{HTML}{98FB98}
\definecolor{TNwhite}{HTML}{FFFFFF}
\definecolor{FNyellow}{HTML}{FFFACD}
\definecolor{FPred}{HTML}{FFB6C1}

\title{\textbf{Vero Cell Culture CPE Detection Benchmarking}}
\author{Albert Mathews \\ Independent Researcher}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This study evaluates the ability of two domain-specific computer-vision tools (AIRVIC and Cellpose) and four general-purpose multimodal large language models (ChatGPT, Claude, Gemini, and Grok) to detect and classify cytopathic effects (CPE) in microscope images of Vero cell cultures. Ground truth CPE presence and morphological types were derived solely from narrative descriptions supplied by the contract research organization (CRO), cross-referenced with standardized descriptors from CLSI M41 Table 7. Quantitative analysis was performed on the 22 images that received CRO descriptions.

None of the evaluated tools demonstrated adequate performance for reliable CPE detection. Domain-specific tools showed extremely high false-positive rates, while the multimodal models exhibited either strong conservative bias or only modest balanced accuracy. These findings indicate that current AI approaches, whether domain-specific or general-purpose, are not yet suitable as objective, reproducible tools for automated CPE analysis in Vero cell culture images.
\end{abstract}

\section{Introduction}
This study benchmarks the performance of two domain-specific computer-vision tools (AIRVIC and Cellpose) and four general-purpose multimodal large language models (ChatGPT, Claude, Gemini, and Grok) on the task of detecting and classifying cytopathic effect (CPE) in Vero-cell culture microscope images. The objective is to evaluate whether any of these AI approaches can serve as objective, reproducible analysis tools for CPE detection in previously unseen images. Ground-truth CPE status and morphological types were derived exclusively from the image descriptions supplied by the contract research organization (CRO) that performed the cell-culture experiments. These descriptions were cross-referenced against the standardized CPE descriptors listed in CLSI M41 Table 7.

\section{Methods}

\subsection{Image dataset and ground truth}
The study utilized the publicly available Vero Cell Culture Image Dataset \cite{mathews2025}. This dataset consists of images acquired in two distinct stages: a preparation (PREP) stage comprising a single path in which cells were thawed and passaged until normal growth rates were achieved, followed by an experimental (EXP) stage. The present study focuses exclusively on the 101 images from the EXP stage. These images were obtained from two experimental paths (Path 1 maintained at 10\% FBS and Path 2 reduced to 2\% FBS) initiated at passage 4. The contract research organization (CRO) was instructed to provide descriptive captions for any image that would benefit from accompanying text; no specific directive was given to search for, classify, or quantify cytopathic effects (CPE). Consequently, only 22 of the 101 EXP-stage images contain CRO descriptions (hereafter referred to as the ``description set''), while the remaining 83 do not. Accuracy analyses were performed exclusively on the \textit{description set}.

\subsection{Standard CPE descriptors}
Characteristic CPE descriptors were extracted from CLSI M41 Table 7 \cite{clsi} through an iterative review process. First, the full table was examined to identify a set of ten descriptors: rounded, ballooned/enlarged, syncytia, vacuolation, detachment, granularity, refractile, cytoplasmic strands, nonspecific degeneration, and no CPE. A CPE descriptor incidence table was then constructed by identifying every instance of these terms (or close semantic correlates) to text entries in the ``Appearance'' column of Table 7. The CLSI M41 Table 7 incidence table is reproduced below for reference.

\input{clsi-table7-descriptors.tex}

Next, the same mapping procedure was applied to the CRO image descriptions to generate a second incidence table (CLSI descriptor set + image descriptions). Only five of the original ten CLSI descriptors appeared in the CRO descriptions. A sixth term, ``Dying Cells'', emerged directly from the CRO text and was retained as a distinct category (Dy) rather than being forced into the CLSI ``nonspecific degeneration'' bin. The resulting \textit{evaluation descriptor set} used for ground truth and all subsequent analyses therefore consisted of: Dying cells (Dy), Rounded (Ro), Vacuolation (V), Detached (D), Granularity (G), and Refractile (Re). The CRO incidence table below is the result of identifying the \textit{evaluation descriptor set} in the image descriptions.

\input{cro-cpe-detections.tex}

\subsection{AIRVIC analysis}
AIRVIC \cite{airvic} is a ResNet-based classifier trained specifically for CPE detection and virus identification in Vero and MDBK cell-culture images. All 101 images of the data set were uploaded to the web interface \cite{arivicweb} with \textit{cell line type} set to ``Vero'' and \textit{virus type} set to ``Unknown''. Because the interface does not support batch export, results were extracted manually and stored in CSV format. AIRVIC outputs a binary CPE decision (present/absent) and a predicted virus identity; the virus predictions are not used in this study, which focuses exclusively on CPE detection.

\subsection{Cellpose analysis}
Cellpose \cite{cellpose} was used as the primary segmentation engine. A custom Python script, developed iteratively in collaboration with Grok, processed all 101 images to generate instance masks---binary images in which each individual cell is delineated as a separate labeled region. Post-processing metrics were then computed on the masks to derive a proxy for CPE presence. Circularity \((4\pi \cdot \text{area} / \text{perimeter}^2)\) was selected because it is a standard morphometric parameter that effectively quantifies the transition from spread polygonal to rounded cell shape, a morphological change widely recognized as a hallmark of cytopathic effect (CPE) in Vero-cell cultures \cite{pmc11180103}. Eccentricity was retained as a complementary descriptor because prior label-free CPE analyses in Vero cultures have identified cell-rounding morphology as a robust, objective visual proxy for cytopathic changes, which can be quantified using shape descriptors such as aspect ratio \cite{revvity}. These two metrics were therefore chosen as the most suitable label-free proxies supported by the literature; the final post-processing pipeline was refined through collaborative testing to ensure the metrics were unaffected by the intentional FBS reduction used in one experimental stage of the dataset.

\input{cellpose-cpe-proxy-metrics.tex}

Binary CPE detection was obtained by converting the combined metrics into a probability score; values $>$ 0.5 were classified as positive. No morphological subtype information was generated. The exact probability formula and post-processing code are available in the project GitHub repository.

\subsection{Multimodal AI analysis}
Four multimodal large language models \cite{llms} were evaluated under identical conditions. For each model a dedicated Python script was written that (1) loads the PNG images, (2) supplies a standardized prompt requesting detection of the six \textit{evaluation descriptor set} morphologies, and (3) requests formatted output. Each model was offered two example images with known ground truth for optional in-context learning. Prompt and processing-script engineering was performed collaboratively with the respective model (i.e. models' are co-authors of the script and prompt). Model outputs were post-processed by a script to produce per-image, per-morphology-type binary decisions that were compared against CRO ground truth.

\subsection{Other tools considered and excluded}
Standard image-processing tools commonly used in cell-culture analysis, as well as the specialized deep-learning tool DVICE, were evaluated but excluded for the following reasons. Fiji/ImageJ \cite{fiji} requires manual filter tuning or macro development that would necessarily incorporate knowledge of the CRO descriptions, introducing circular reasoning and violating the requirement for an objective, a-priori tool. CellProfiler \cite{cellprofiler} was tested with the ``ExampleHuman'' pipeline; however, the pipeline proved too rigid for the present image set, and any custom pipeline development would again risk circular reasoning. Snapcyte \cite{snapcyte} could not be accessed after repeated failed account-creation attempts and lack of response from the vendor. Revvity Signals Image Artist \cite{revvityIA} was discussed with the manufacturer but commercial and technical setup barriers prevented its inclusion. For DVICE \cite{pmc11180103}, contact was made with the authors requesting access to the code or a means to use the tool; a response is still pending. These exclusions are reported for transparency; the tools ultimately benchmarked represent the most readily available and least biased options that could be applied without post-hoc tuning to the ground truth.

\subsection{Statistical evaluation}
Performance was quantified using false-positive rate, false-negative rate, true-positive rate, true-negative rate, and overall accuracy. For the multimodal models these metrics were first computed separately for each of the six CPE morphological types and then macro-averaged. For AIRVIC and Cellpose the metrics were calculated directly on the binary CPE detection task. Baseline ``always-positive'' and ``always-negative'' classifiers were also evaluated for comparison. All calculations were performed on the 22-image \textit{description set} only.

\section{Results}
Performance results are presented the tables below.
\input{airvic-results.tex}
\input{cellpose-results.tex}
\input{llm-cpe-detections.tex}
\subsection{Aggregate performance summary}
Performance metrics of AIRVIC, CellPose, and four large language models for cytopathic effect (CPE) detection are shown in Table 8.
AIRVIC and CellPose are compared against simple always true and always false binary classification models.
LLMs are compared against always true and always false CPE type classification models. The accuracy rate of the \textit{always false CPE type classification} model is a characteristic of the dataset. For any model to prove useful, it would have to significantly outperform the \textit{always false CPE type classification} model.

\input{aggregate-results.tex}

\section{Discussion}

\subsection{Domain-specific tools}
AIRVIC detected CPE in 21 of the 22 \textit{description set} images and in 100 of the 101 images of the EXP-stage, producing a high false-positive rate (0.909) and overall accuracy of only 0.545. The test images are visually similar to the Vero-cell images used in AIRVIC's original training set, as demonstrated by the table below. Despite this similarity, the model failed to generalize effectively to the present dataset.
Cellpose similarly classified nearly all images as CPE-positive (accuracy 0.500), again driven by an extremely high false-positive rate. The morphological-proxy approach therefore did not provide useful discrimination in this dataset.

\input{image-processing-vs-airvic-expectation-table.tex}

\subsection{Multimodal AI tools}
The four general-purpose multimodal models exhibited macro-averaged accuracies between 0.542 and 0.709. Gemini achieved the highest overall accuracy (0.709) but at the cost of a very high false-negative rate, effectively behaving like a conservative ``no-CPE'' classifier; this performance was still lower than that of the simple always-negative baseline model. ChatGPT, Claude, and Grok showed poor performance (accuracies 0.549--0.559) with substantial rates of both false positives and false negatives. None of the multimodal models reached a level of reliability that would support their use as objective CPE detection tools for this dataset.

\subsection{Limitations}
The ground-truth labels are based on narrative descriptions provided by the contract research organization (CRO) that were not generated under a CPE-specific protocol. The description set is small (\(n = 22\)), and the remaining 83 images could not be evaluated quantitatively. Domain-specific tools were tested exactly as publicly available, with no retraining or fine-tuning performed. Prompt engineering for the multimodal models, although extensive, remains inherently model-dependent.

Furthermore, the images from the experimental stage exhibited relatively low cell confluency compared with typical culture conditions under which CPE is normally observed. It is possible that the training sets of the domain-specific models did not include images with similarly low confluency.

\subsection{Implications}
None of the evaluated AI tools demonstrated sufficient accuracy or robustness to replace or act as an objective second opinion for human expert interpretation of CPE in Vero-cell cultures under the conditions tested. Future work may benefit from larger, prospectively labeled datasets and from domain-specific fine-tuning of multimodal models.

\begin{thebibliography}{10}

\bibitem{mathews2025} Mathews, A. (2025). Vero Cell Culture Image Dataset [Data set]. Zenodo. https://doi.org/10.5281/zenodo.17928456

\bibitem{clsi} CLSI M41. Viral Culture. Clinical and Laboratory Standards Institute. https://clsi.org/shop/standards/m41/

\bibitem{airvic} Akkutay-Yoldar, Z., Yoldar, M.T., Akkaş, Y.B. et al. (2025). A web-based artificial intelligence system for label-free virus classification and detection of cytopathic effects. \textit{Sci Rep} 15, 5904. https://doi.org/10.1038/s41598-025-89639-0

\bibitem{airvicweb} AIRVIC web interface. https://airvic.turkai.com/ (accessed 16 April 2026).

\bibitem{cellpose} Stringer, C. et al. (2021). Cellpose: a generalist algorithm for cellular segmentation. \textit{Nature Methods}.

\bibitem{pmc11180103} PMC11180103. Vero-cell CPE quantification reference. https://pmc.ncbi.nlm.nih.gov/articles/PMC11180103/

\bibitem{revvity} Revvity. Cytopathic effects analysis notes. https://www.revvity.com/ask/cytopathic-effects

\bibitem{llms} ChatGPT (OpenAI), Claude (Anthropic), Gemini (Google), Grok (xAI) -- model versions used in 2025--2026.

\bibitem{fiji} Schindelin, J., Arganda-Carreras, I., Frise, E. et al. (2012) Fiji: an open-source platform for biological-image analysis. \textit{Nature Methods} \textbf{9}, 676–682.

\bibitem{cellprofiler} Carpenter, A.E., Jones, T.R., Lamprecht, M.R. et al. (2006) CellProfiler: image analysis software for identifying and quantifying cell phenotypes. \textit{Genome Biology} \textbf{7}, R100.

\bibitem{snapcyte} SnapCyte. AI Image Analysis \& Custom AI Models. https://www.snapcyte.com/ (accessed 16 April 2026).

\bibitem{revvityIA} Revvity. Signals Image Artist. https://www.revvity.com/ca-en/product/signals-image-artist-sw-5-user-subscript-inf02358 (accessed 16 April 2026).

\end{thebibliography}

\end{document}